# The relative age effect in professional biathlon
This Rmarkdown notebook is a piece of a more broad work that sees the biggest part of the data processing part developed in Python.
The goal is to inspect the relative age effect (RAE) in professional biathlon. A Poisson one-sided hypothesis test is performed on a group of selected nations and a further inspection is done on the most competitive nations.






```{r}
library(ggplot2)
library(rjson)
library(sfsmisc)
library(coda)
library(rjags)
```



```{r}
# Importing the Json data into a list
json_data <- fromJSON(file='Rdata.json')
```




```{r}
# Declaring the variables mean and standard deviation for the priors
mean_early <- json_data$data_num$mean$early
mean_late <- json_data$data_num$mean$late
std_early <-json_data$data_num$std$early
std_late <-json_data$data_num$std$late
```


```{r}
# Declaring the variables mean and standard deviation for the priors of the elite nations
mean_early_nor <- json_data$data_nor$mean$early
mean_late_nor <- json_data$data_nor$mean$late
std_early_nor <-json_data$data_nor$std$early
std_late_nor <-json_data$data_nor$std$late
```

```{r}
# Defining the functions to map mean and std to the parameters of the beta
alpha_gamma <- function(mean, std){
  return( (mean^2)/(std^2) )
}

beta_gamma <- function(mean, std){
  return( (mean)/(std) )
}
```



```{r}
# Finding the parameters of the beta
alpha_prior_early <- alpha_gamma(mean_early, std_early)
alpha_prior_late <- alpha_gamma(mean_late, std_late)
alpha_prior_early_nor <- alpha_gamma(mean_early_nor, std_early_nor)
alpha_prior_late_nor <- alpha_gamma(mean_late_nor, std_late_nor)
```


```{r}
# Finding the parameters of the beta
beta_prior_early <- beta_gamma(mean_early, std_early)
beta_prior_late <- beta_gamma(mean_late, std_late)
beta_prior_early_nor <- beta_gamma(mean_early_nor, std_early_nor)
beta_prior_late_nor <- beta_gamma(mean_late_nor, std_late_nor)
```

```{r}
# Retrieving data for the inference
y_early <- json_data$first_six$first_six_num
y_late <- json_data$second_six$second_six_num

y_early_nor <- json_data$first_six$first_six_nor
y_late_nor <- json_data$second_six$second_six_nor
```


```{r}
# Updating the parameters for the inference
alpha_post_early <- alpha_prior_early + y_early
alpha_post_late <- alpha_prior_late + y_late

alpha_post_early_nor <- alpha_prior_early_nor + y_early_nor
alpha_post_late_nor <- alpha_prior_late_nor + y_late_nor
```

```{r}
# Setting the random seed
set.seed(123)
```


```{r}
# Updating the parameters for the inference
n <- json_data$tot_years


beta_post_early <- beta_prior_early + n
beta_post_late <- beta_prior_late + n

beta_post_early_nor <- beta_prior_early_nor + n
beta_post_late_nor <- beta_prior_late_nor + n
```




```{r}
options(repr.plot.width=15, repr.plot.width=7)
x_axis_1 = seq(0, 9, length.out=1000)
plot <- ggplot() +
  geom_line(aes(x = x_axis_1, y = dgamma(x = x_axis_1, shape = alpha_post_early, 
                                         rate = beta_post_early), 
                                        color='Early borns'), lwd=1.5) +
  geom_line(aes(x = x_axis_1, y = dgamma(x = x_axis_1, shape = alpha_post_late, rate = beta_post_late),
                                        color='Late borns'), lwd=1.5) +
  ylim(0,1.3) +
  labs(color='Legend', title = 'Poisson distribution of early and late borns', y='PDF', 
       x='Events per year') +
  theme_bw()

plot
```

```{r}
mean_f <- alpha_post_early/beta_post_early - alpha_post_late/beta_post_late
std_f <- sqrt((alpha_post_early/beta_post_early^2)^2 + (alpha_post_late/beta_post_late^2)^2)


mean_f_nor <- alpha_post_early_nor/beta_post_early_nor - alpha_post_late_nor/beta_post_late_nor
std_f_nor <- sqrt((alpha_post_early_nor/(beta_post_early_nor^2))^2+(
  alpha_post_late_nor/(beta_post_late^2))^2)
```


## Numerical arroximation
Due to the fact that the difference between two gamma distributions is quite the complicated issue, we rely to numerical approximations to perform the hypothesis testing.
We in fact plot the histogram of the difference between the two distribution by taking the difference between two sampled sets. We then integrate the density approximation of the curve after normalizing it: the density estimate has indeed a numerical integral that is slightly removed from 1.


 
```{r}
# Creating the two sampled sets
sim_early <- rgamma(n = 10000, shape = alpha_post_early, 
                                         rate = beta_post_early)
sim_late <- rgamma(n = 10000, shape = alpha_post_late, 
                                         rate = beta_post_late)

sim_diff = sim_early - sim_late


sim_early_nor <- rgamma(n = 10000, shape = alpha_post_early_nor, 
                                         rate = beta_post_early_nor)
sim_late_nor <- rgamma(n = 10000, shape = alpha_post_late_nor, 
                                         rate = beta_post_late_nor)

sim_diff_nor = sim_early_nor - sim_late_nor


```


```{r}
# Defining the Freedman-Diaconis rule in order to apply it for the bin's choice
bins_fd <- function(vec) {
  diff(range(vec)) / (2 * IQR(vec) / length(vec)^(1/3))
}
```



```{r}
# Numerical integration
density_est_num <-  density(sim_diff)
lower_dens_x_num <- density_est_num$x[density_est_num$x < 0]
lower_dens_y_num <- density_est_num$y[density_est_num$x < 0]
integral_num <- integrate.xy(x = lower_dens_x_num, fx = lower_dens_y_num) 

integral_num
```

```{r}
# Finding the right normalization factor
norm_int = integrate.xy(x = density_est_num$x, fx=density_est_num$y)
norm_int
```



```{r}
x_axis = seq(-10, 10, length.out=1000)
x_ribbon <- seq(-10, 0, length.out=1000)



plot <- ggplot() +
    geom_histogram((aes(x = sim_diff, y=..density..)), bins=bins_fd(sim_diff), color='white',
                    fill='navy',alpha=0.8)+
  labs(title = 'Differenza tra nati nel 1° e nel 4° quadrimestre', y='PDF', x='Differenza del numero di eventi annuali') +
  #geom_line(aes(x = x_axis, y = dnorm(x = x_axis, mean = mean_f, sd = std_f)), color='firebrick', 
   #         lwd = 1.5 ) +
  geom_line(aes(x = density_est_num$x, y=density_est_num$y/norm_int), color='firebrick', lwd=1.5) +

  geom_ribbon(aes(x = lower_dens_x_num, ymin=0, ymax=lower_dens_y_num), 
              fill='cyan2', color='black', alpha=0.5, lwd=1)+
  geom_segment(aes(x = 0, xend=0, y=0, yend=lower_dens_y_num), lwd=1) +
  theme_bw() +
  annotate('text', x=-2.3, y=0.1, label= sprintf('Valore dell\'integrale = %s', round(integral_num, 3))) +
   geom_segment(aes(x = -0.3, y = 0.1, xend = -1.3, yend = 0.1),
                  arrow = arrow(length = unit(0.3, "cm"))) +
  
  scale_x_continuous(limits = c(-3, 2.5))
  

ylim(0,0.35) 

plot 

```

## Hypothesis test results
We can see how, setting the significance level to 10% we can reject the null hypothesis of late born being more present in the word cup than early born. Our conclusion is that with 90% probability an effect shifting the distribution toward athletes born in the first trimester can be detected with our hypothesis testing. 





```{r}
# Numerical integration for the elite nations
density_est <-  density(sim_diff_nor)
lower_dens_x <- density_est$x[density_est$x < 0]
lower_dens_y <- density_est$y[density_est$x < 0]
integral_nor <- integrate.xy(lower_dens_x, lower_dens_y) 

integral_nor
```





```{r}
plot <- ggplot() +
        geom_histogram(aes(x= sim_diff_nor, y=..density..), bins = bins_fd(sim_diff_nor), 
                       fill='navy', color='white', alpha=0.8) +
        geom_line(aes(x= density_est$x, y= density_est$y), lwd=1.5, color='firebrick') +
        geom_ribbon(aes(x = lower_dens_x, ymin=0, ymax=lower_dens_y), 
                    fill='cyan2', color='black', alpha=0.5, lwd=1)+
        annotate('text', x=-0.5, y=0.5, label= sprintf('Integral value = %s', round(integral_nor, 3))) +
        geom_segment(aes(x = -0.2, y = 0.07, xend = -0.3, yend = 0.45),
                  arrow = arrow(length = unit(0.3, "cm"))) +
        geom_segment(aes(x = 0, xend=0, y=0, yend=lower_dens_y), lwd=1) +
        labs(title = 'Difference between late and early borns in Norway', 
             x='Difference in the number of events', y = 'PDF') +
        theme_bw() 
plot
```

## Second hypothesis test results
Following the hypothesis of the RAE effect, numerous research papers conclude that the effect is more present in more competitive frameworks.
We have here analyzed the 3-most competitive nations in terms of number of athletes with at least one berth in a word-cup race. The percentage of early borns is higher when considering this elite triplet of nations. Due to the low number of samples though this result cannot be considered statistically significant.
This results suggest a more in-depth investigation of RAE, with a focus that should also be put to junior competitors. A focus that treats separately some previously selected nations could be beneficial to inspect how differences in the handling of athlete's development affect the presence of the RAE or the lack of it.




























